---
title: "data_wrangling"
author: "Francis"
date: "11/10/2018"
output: github_document
editor_options: 
  chunk_output_type: console
---

After consideration, the `admissions.csv` database seems really interesting. It is useful to analyze the  connection between mutiple factors and death.
```{r}
library(tidyverse)
library(devtools)
library(readr)
library(lubridate)
library(broom)


```

# MIMIC3

### Import data
```{r}
admissions <- 
  read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names()
names(admissions)
  
# The year should be delt with to become normal. The discharge time means the time when the patient leave the hospital.


# see types
class(admissions$admittime)
class(admissions$dischtime)
class(admissions$deathtime)
class(admissions$admission_type) 
class(admissions$insurance)
class(admissions$religion)
class(admissions$ethnicity)
class(admissions$edregtime)
class(admissions$edouttime)

```
All of them are factors.



### Tidy data
```{r}
#Create year, month, day variables
admissions <- 
  admissions %>% 
  separate(admittime, into = c("admittime_year", "admittime_month", "admittime_day"), sep = "-")
admissions <- 
  admissions %>% 
  separate(dischtime, into = c("dischtime_year", "dischtime_month", "dischtime_day"), sep = "-") %>% 
  separate(dischtime_day, into = c("dischtime_day", "dischtime_time"), sep = " ")
admissions <- 
  admissions %>% 
  separate(deathtime, into = c("deathtime_year", "deathtime_month", "deathtime_day"), sep = "-") %>% 
  separate(deathtime_day, into = c("deathtime_day", "deathtime_time"), sep = " ")
admissions <- 
  admissions %>% 
  separate(edregtime, into = c("edregtime_year", "edregtime_month", "edregtime_day"), sep = "-") %>% 
  separate(edregtime_day, into = c("edregtime_day", "edregtime_time"), sep = " ")
admissions <- 
  admissions %>% 
  separate(edouttime, into = c("edouttime_year", "edouttime_month", "edouttime_day"), sep = "-") %>% 
  separate(edouttime_day, into = c("edouttime_day", "edouttime_time"), sep = " ")

# Correct year to normal
admissions <- 
  admissions %>% 
  mutate(admittime_year = as.numeric(admittime_year) - 200, dischtime_year = as.numeric(dischtime_year) - 200, deathtime_year = as.numeric(deathtime_year) - 200, edregtime_year = as.numeric(edregtime_year) - 200, edouttime_year = as.numeric(edouttime_year) - 200)
# Take a look
head(admissions)
```





### linear regression

```{r}
# read original data
admissions_origin <- 
  read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names()


# add a death factor and duration factor
# exclude NEWBORN, because they are not treated as patients.
### There are too many diagnosis, so the diagnosis will be analyzed by some analysis of scores in another part.
admissions_death <- 
  mutate(admissions_origin, living = is.na(admissions_origin$deathtime), hospitaltime =  admissions_origin$dischtime - admissions_origin$admittime, edtime = admissions_origin$edouttime - admissions_origin$edregtime) %>% 
  filter(diagnosis != "NEWBORN")

```



```{r}
# glance data
skimr::skim(admissions_death)

```



```{r}


### try logistic regression step by step.

#SLR

par(mfrow = c(2,2))
living_lm1 <- 
  glm(living ~ admission_type, data = admissions_death) %>% 
  plot()
summary(living_lm1)
broom::glance(living_lm1)
broom::tidy(living_lm1)

living_lm2 <- 
  glm(living ~ admission_location, data = admissions_death) %>% 
  plot()
summary(living_lm1)
broom::glance(living_lm2)
broom::tidy(living_lm2)

living_lm3 <- 
  glm(living ~ insurance, data = admissions_death) %>% 
  plot()
summary(living_lm3)
broom::glance(living_lm3)
broom::tidy(living_lm3)

living_lm4 <- 
  glm(living ~ language, data = na.omit(select(admissions_death, living, language))) %>% 
  plot()
summary(living_lm4)
broom::glance(living_lm4)
broom::tidy(living_lm4)

living_lm5 <- 
  glm(living ~ religion, data = admissions_death) %>% 
  plot()
summary(living_lm5)
broom::glance(living_lm5)
broom::tidy(living_lm5)

living_lm6 <- 
  glm(living ~ marital_status, data = admissions_death) %>% 
  plot()
summary(living_lm6)
broom::glance(living_lm6)
broom::tidy(living_lm6)

living_lm7 <- 
  glm(living ~ ethnicity, data = admissions_death) %>% 
  plot()
summary(living_lm7)
broom::glance(living_lm7)
broom::tidy(living_lm7)

#living_lm8 <-   Too many diagnosis 
#  glm(living ~ diagnosis, data = admissions_death)
#summary(living_lm8)

living_lm9 <- 
  glm(living ~ hospital_expire_flag, data = admissions_death) %>% 
  plot()
summary(living_lm9)
broom::glance(living_lm9)
broom::tidy(living_lm9)

living_lm10 <- 
  glm(living ~ has_chartevents_data, data = admissions_death) %>% 
  plot()
summary(living_lm10)
broom::glance(living_lm10)
broom::tidy(living_lm10)

living_lm11 <- 
  glm(living ~ hospitaltime, data = admissions_death) %>% 
  plot()
summary(living_lm11)
broom::glance(living_lm11)
broom::tidy(living_lm11)

living_lm12 <- 
  glm(living ~ edtime, data = na.omit(select(admissions_death, living, edtime))) %>% 
  plot()
summary(living_lm12)
broom::glance(living_lm12)
broom::tidy(living_lm12)
```



```{r eval=FALSE}
#MLR
living_mlr <- 
  glm(living ~ admission_type + admission_location + insurance + language + religion + marital_status + ethnicity + hospital_expire_flag + has_chartevents_data + hospitaltime + edtime, data = admissions_death)
broom::glance(living_mlr)







# Do more regression analysis

coefficients(living_mlr)
confint(living_mlr)
fitted(living_mlr)
residuals(living_mlr)
anova(living_mlr)
vcov(living_mlr)


###backward elimination
#Backward elimination: take out non-significant variables ’one at a time’ starting with the highest p-value
step1 <- update(living_mlr, . ~. -admission_type)
broom::glance(step1)
# remove

step2 <- update(living_mlr, . ~. -admission_type-admission_location)
broom::glance(step2)
#keep

step3 <- update(living_mlr, . ~. -admission_type-insurance)
broom::glance(step3)
#keep

step4 <- update(living_mlr, . ~. -admission_type-language)
broom::glance(step4)
#remove

step5 <- update(living_mlr, . ~. -admission_type-language-religion)
broom::glance(step5)
#keep

step6 <- update(living_mlr, . ~. -admission_type-language-marital_status)
broom::glance(step6)
#keep

step7 <- update(living_mlr, . ~. -admission_type-language-ethnicity)
broom::glance(step7)
#keep

step8 <- update(living_mlr, . ~. -admission_type-language-hospital_expire_flag)
broom::glance(step8)
#keep

step9 <- update(living_mlr, . ~. -admission_type-language-has_chartevents_data)
broom::glance(step9)
#keep

step10 <- update(living_mlr, . ~.  -admission_type-language-hospitaltime)
broom::glance(step10)
#keep

step11 <- update(living_mlr, . ~. -admission_type-language-edtime)
broom::glance(step11)
#keep

```

### So we got this regression model:



\item $\hat{Living}$ = \item $\hat{\beta_0}$ + \item $\hat{\beta_1}$ Admission.Location + \item $\hat{\beta_2}$ Insurance + \item $\hat{\beta_3}$ Insurance + \item $\hat{\beta_4}$ Religion + \item $\hat{\beta_5}$ Marital.Status + \item $\hat{\beta_6}$ Ethnicity + \item $\hat{\beta_7}$ Hospital.Expire.Flag + \item $\hat{\beta_8}$ Has.ChartEvents.Data + \item $\hat{\beta_9}$ Hospital.Time + \item $\hat{\beta_10}$ Edtime 






















Maybe we need this step:
```{r}
# Remove NA lines
#admissions_death_nona <- 
#  admissions_death %>% 
#  select(admission_type, admission_location, insurance, religion, marital_status, ethnicity, hospitaltime, edtime) %>% 
#  na.omit() 

#MLR with no NAs
#living_mlr_nona <- 
#  glm(living ~ admission_type + admission_location + insurance + language + religion + marital_status + ethnicity + hospital_expire_flag + has_chartevents_data + hospitaltime + edtime, data = admissions_death_nona)
#summary(living_mlr_nona)
#broom::glance(living_mlr_nona)
#broom::tidy(living_mlr_nona)
```





























#OpenFDA

### Import OpenFDA
```{r}
# Already installed openfda data
# Load OpenFDA
library(openfda)


library(jsonlite)
fda <- 
  fromJSON("https://api.fda.gov/drug/event.json") %>% 
  janitor::clean_names()

names(fda$results)

```
There are `r length(fda$results)` observations in the `event` dataset under `drug`, they are: `r names(fda$results)`.



















### Visulization
Let's take a look at the combination of MIMIC data and ICU together(for we want to know the ICU time since it is more critical for mortality)

```{r}
icu <- 
  read_csv("./database/data/icu_detail.csv") %>% 
  janitor::clean_names()

# See what is the types of columns in icu dataset
output = vector("list", length = 19)

for (i in 1:19) {
  output[[i]] = class(icu[[i]])
}
```




Let's combine these two dataset so we can visualize it correctly

```{r}

total_patient <-
  merge(admissions_death, icu, by="hadm_id") %>% 
  as.tibble()

#mutate icu time
total_patient <- 
  mutate(total_patient, icutime = outtime - intime)

skimr::skim(total_patient) %>% 
  knitr::kable()

head(total_patient) %>% 
  knitr::kable()


```




Let's see some plots from these two datasets

```{r}
hospitaltimeplot <- ggplot(data = total_patient, aes(x = diagnosis, y = hospitaltime)) + 
geom_histogram()

icutimeplot <- ggplot(data = total_patient, aes(x = diagnosis, y = icutime)) + 
geom_histogram()
```


Let's focus on the `hospitaltime` and `icutime`


```{r eval=FALSE}
mlr_time <- 
  glm(living ~ hospitaltime + icutime, data = total_patient)
plot(mlr_time)

plot(total_patient)


```

