---
title: "icu details"
author: "Laura Cosgrove"
date: "12/1/2018"
output: github_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(RPostgreSQL)
library(tidyverse)
library(dbplyr)
```

The purpose of running this database query is to gain a clean table of demographic data for all patients who were in the ICU. We are choosing SQL for this query because of the comparative simplicity of working with time data in SQL.

## Database

```{r dbconnect}
# Load configuration settings
dbdriver <- 'PostgreSQL'
host  <- '127.0.0.1'
port  <- '5432'
user  <- 'postgres'
password <- 'postgres'
dbname <- 'mimic'
schema <- 'mimiciii'
# Connect to the database using the configuration settings
con <- dbConnect(dbDriver(dbdriver), dbname = dbname, host = host, port = port, 
                 user = user, password = password)
# Set the default schema
dbExecute(con, paste("SET search_path TO ", schema, sep=" "))
```

Set this database as the connection for all future sql chunks:

```{r}
knitr::opts_chunk$set(connection = "con")
```

```{sql connection=con, eval = FALSE}
-- ------------------------------------------------------------------
-- Title: Detailed information on ICUSTAY_ID
-- Description: This query provides a useful set of information regarding patient
--              ICU stays. The information is combined from the admissions, patients, and
--              icustays tables. It includes age, length of stay, sequence, and expiry flags.
-- MIMIC version: MIMIC-III v1.3
-- ------------------------------------------------------------------

-- (Optional) Define which schema to work on
SET search_path TO mimiciii;

-- This query extracts useful demographic/administrative information for patient ICU stays
DROP MATERIALIZED VIEW IF EXISTS icustay_detail CASCADE;
CREATE MATERIALIZED VIEW icustay_detail as

SELECT ie.subject_id, ie.hadm_id, ie.icustay_id

-- patient level factors
, pat.gender, pat.dod

-- hospital level factors
, adm.admittime, adm.dischtime
, ROUND( (CAST(EXTRACT(epoch FROM adm.dischtime - adm.admittime)/(60*60*24) AS numeric)), 4) AS los_hospital
, ROUND( (CAST(EXTRACT(epoch FROM adm.admittime - pat.dob)/(60*60*24*365.242) AS numeric)), 4) AS admission_age
, adm.ethnicity, adm.admission_type
, adm.hospital_expire_flag
, DENSE_RANK() OVER (PARTITION BY adm.subject_id ORDER BY adm.admittime) AS hospstay_seq
, CASE
    WHEN DENSE_RANK() OVER (PARTITION BY adm.subject_id ORDER BY adm.admittime) = 1 THEN True
    ELSE False END AS first_hosp_stay

-- icu level factors
, ie.intime, ie.outtime
, ROUND( (CAST(EXTRACT(epoch FROM ie.outtime - ie.intime)/(60*60*24) AS numeric)), 4) AS los_icu
, DENSE_RANK() OVER (PARTITION BY ie.hadm_id ORDER BY ie.intime) AS icustay_seq

-- first ICU stay *for the current hospitalization*
, CASE
    WHEN DENSE_RANK() OVER (PARTITION BY ie.hadm_id ORDER BY ie.intime) = 1 THEN True
    ELSE False END AS first_icu_stay

FROM icustays ie
INNER JOIN admissions adm
    ON ie.hadm_id = adm.hadm_id
INNER JOIN patients pat
    ON ie.subject_id = pat.subject_id
WHERE adm.has_chartevents_data = 1
ORDER BY ie.subject_id, adm.admittime, ie.intime;
```


Set evaluation to false so that you don't requery your database every time you knit. Query from saved file:
```{r}
icu_detail_view <- read_file("./database/mimic-code/concepts/demographics/icustay-detail.sql")


#Generate materialized views
dbGetQuery(con, icu_detail_view)
```

Select data from the ICU detail:

```{r}
icu_detail_query <- "SELECT *
              FROM icustay_detail i;"
icu_detail_data <- as.tibble(dbGetQuery(con, icu_detail_query))
write_csv(icu_detail_data, path = "./database/icu_detail.csv")
icu_detail_data
```

The ICU detail query extracts both hospital level length of stay as well as ICU-level length of stay, in terms of days.

