---
title: "Diagnostic Categories"
author: "Laura Cosgrove"
date: "12/4/2018"
output: github_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(RPostgreSQL)
library(tidyverse)
library(dbplyr)
```

This R code makes use of the Clinical Classification Software (CCS), which categorizes ICD-9 coded diagnoses into clinically meaningful groups. The categorization was developed by the Agency for Healthcare Research and Quality (AHRQ). More detail can be found on the AHRQ website: https://www.hcup-us.ahrq.gov/tools_software.jsp.

This software contains two tables: `ccs_single_level_dx.csv.gz` and `ccs_multi_level_dx.csv.gz`. The first table contains high-level diagnostic category infomation, while the multi-level diagnostic category information contains four levels of diagnostic detail.

The purpose of reading in AHRQ's clinical categories is to help make diagnosis descriptions for our data more intuitive.

##Read in data

```{r}
multi_ccs <- read_csv("./database/mimic-code/concepts/diagnosis/ccs_multi_level.csv.gz",
                      col_names = c("ccs_cat_multi", "description_multi_1", "description_multi_2", "description_multi_3", "description_multi_4","icd9_code"))
single_ccs <- read_csv("./database/mimic-code/concepts/diagnosis/ccs_single_level.csv.gz",
                       col_names = c("ccs_cat_single", "description_single", "icd9_code"))
d_icd_diagnoses <- read_csv("./database/data/D_ICD_DIAGNOSES.csv.gz") %>% 
  janitor::clean_names()
```

### Inner join

```{r}
d_icd_diagnoses_single <- d_icd_diagnoses %>% 
  inner_join(., single_ccs, by = "icd9_code") 

d_icd_diagnoses_multi <- d_icd_diagnoses %>% 
  inner_join(., multi_ccs, by = "icd9_code") 

```

Now, the diagnosis coding the hospital uses is given overall diagnostic categories.

## Join with the patient data

```{r}
diagnoses_icd <- read_csv("./database/data/DIAGNOSES_ICD.csv.gz") %>% 
  janitor::clean_names()

diagnoses_icd_single_cat <- diagnoses_icd %>% 
  inner_join(., d_icd_diagnoses_single, by = "icd9_code") 

#Our high-level categories
diagnoses_icd_single_cat %>% 
  distinct(description_single)

#The EMR high-level categories
diagnoses_icd_single_cat %>% 
  distinct(short_title)

#Unique ICD 9 codes: about the same dimension as the short titles for diagnoses
diagnoses_icd_single_cat %>% 
  distinct(icd9_code)

diagnoses_icd_single_cat %>% 
  distinct(short_title)

#Another comparision, initial diagnosis from admissions dataset: > 15,000 unique
admissions <- read_csv("./database/data/ADMISSIONS.csv.gz") %>% 
  janitor::clean_names()

admissions %>% 
  distinct(diagnosis)

# Check if the sheet succesfully joins to admissions. Should be multiple diagnoses per subject
admissions %>% 
  inner_join(., diagnoses_icd_single_cat, by = "hadm_id") %>% 
  select(hadm_id, diagnosis, icd9_code, short_title, description_single) %>% 
  group_by(hadm_id)

```

We've succesfully reduced the complexity of the high-level diagnostic categories. We may be missing some important data, but we could try to remedy that, in part, by joining the multi_ccs dataset to our patient data. But for now, I will export the short datasheet of patients + diagnosis category so that my team members can join it to admissions data for their analyses.

## Export diagnosis with subject data and admissions diagnosis

```{r}
subject_diag_cat <- admissions %>% 
  inner_join(., diagnoses_icd_single_cat, by = "hadm_id") %>% 
  select(hadm_id, hospital_expire_flag, diagnosis, icd9_code, short_title, description_single) %>% 
  rename(death_bin = hospital_expire_flag, admit_diagnosis = diagnosis, category = description_single) %>% 
  group_by(hadm_id)

write_csv(subject_diag_cat, "./database/subject_diag_cat.csv")

 
```

