---
title: "length_of_care"
author: "Samantha Brown"
date: "11/17/2018"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(devtools)
library(readr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
admissions <- 
  read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names() %>% 
  mutate(diagnosis = factor(diagnosis))
```

## Exploratory analysis

```{r, output = FALSE}
n_admits = admissions %>% 
  nrow()

n_discharge = admissions %>% 
  filter(is.na(deathtime)) %>% 
  nrow()

round((1 - (n_discharge/n_admits)) * 100, digits = 2)
```

**Out of `r admissions %>% nrow()` patient admissions, `r admissions %>%  filter(is.na(deathtime)) %>%  nrow()` patients were ultimately discharged. The remaining `r round((1 - (n_discharge/n_admits)) * 100, digits = 2)`% were recorded deaths.**


```{r}
## Ethnicity
ethnicity = admissions %>% 
  group_by(ethnicity) %>% 
  count()
```

Need to do string replace to combine similar groups.

```{r}
## Marital Status
marital_status = admissions %>% 
  group_by(marital_status) %>% 
  count() %>% 
  mutate(percent = n/nrow(admissions)*100) %>% 
  arrange(desc(percent))
```

`r marital_status[1,3]`% of the critical care patients were married, while `r marital_status[2, 3]`% of patients were single. The remaining `r round(100 - (marital_status[1,3] + marital_status[2, 3]), digits = 2)`% were either widowed, divorced, separated, had a life partner, or marital status was unknown.

```{r}
## Distribution of insurance type
admissions %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(percent_of_patients = round(n/nrow(admissions)*100, digits = 2)) %>% 
  arrange(desc(percent_of_patients)) %>% 
  select(-n)
```



```{r}
## Top 10 causes of recorded mortalities
admissions %>% 
  filter(deathtime > 0) %>%
  count(diagnosis) %>% 
  top_n(10) %>% 
  arrange(desc(n))
```

```{r}
## Look at sepsis diagnoses
sepsis = admissions %>% 
  filter(diagnosis == "SEPSIS") %>% 
  ## discharge = 1, mortality = 0
  mutate(mortality = ifelse(is.na(deathtime), 1, 0)) %>% 
  ## create indicator variable: married = 1, not married = 0
  mutate(marital_status = ifelse(marital_status == "MARRIED", 1, 0)) %>% 
  select(subject_id, insurance, marital_status, ethnicity, mortality)


## Look at top 10 causes of mortalities
top_10_causes = admissions %>% 
  filter(diagnosis == c("SEPSIS", "PNEUMONIA", "INTRACRANIAL HEMORRHAGE", "CONGESTIVE HEART FAILURE", "ALTERED MENTAL STATUS", "CARDIAC ARREST", "ABDOMINAL PAIN", "S/P FALL", "HYPOTENSION", "SUBARACHNOID HEMORRHAGE"))
```



```{r}
## Insurance -- mortality vs. no mortality 

## mortality
mortalities = admissions %>% 
  filter(deathtime > 0) %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(outcome = "mortality")

total_mortalities = sum(mortalities$n)

mortalities = mortalities %>% 
  mutate(patient_proportion = round(n/total_mortalities, digits = 4))

discharge = admissions %>% 
  filter(deathtime > 0) %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(outcome = "no_mortality")

total_discharge = sum(discharge$n)

dishcharge = discharge %>% 
  mutate(patient_proportion = round(n/total_discharge, digits = 4)) %>% 
  rbind(mortalities) %>% 
  mutate(log_count = log(n)) %>% 
  ggplot(aes(x = reorder(insurance, log_count), y = log_count)) +
  geom_point(color = "blue") 
  
```


## Exploratory:
I want to discuss with you guys the challenge in finding length of care -- if someone was admitted on December 1, 1980 and discharged January 5, 1981, I'm not sure if we can just subtract the times? Using Francis's method, it results in negative number for month. Also, we cannot measure minutes/hours, because admit time does not have minutes

Should we measure length of stays by hours? then we can combine day and hour to one unit measurement

```{r}
library(lubridate)
admissions_data <- read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names()

```

```{r}
difference = admissions_data %>% 
  mutate(length_of_stay =  lubridate::as.duration(admittime %--% dischtime),
         mortality = ifelse(is.na(deathtime), 1, 0))

summary(difference$length_of_stay)

stay_in_days = difference %>% 
  mutate(length_of_stay = as.numeric(length_of_stay)/(60*60*24))


max(difference$length_of_stay)


max_length_in_days = max(as.numeric(difference$length_of_stay)/(60*60*24))
-1*min(as.numeric(difference$length_of_stay))/(60*60*24)

```


```{r}
## Merge ICU data with admissions dataset 
## Now we have length of hospital stay and length of ICU stay
icu_data = read_csv("./database/data/icu_detail.csv")

icu_data = icu_data %>% 
  select(hadm_id, los_icu, los_hospital, icustay_seq, first_icu_stay)

merged_data = left_join(difference, icu_data, by = "hadm_id")

class(merged_data$los_icu)
class(merged_data$length_of_stay)
```


```{r}
ggplot(merged_data, aes(x = los_icu, y = log(as.numeric(length_of_stay)))) +
  geom_point() +
  labs(
    x = "Length of Stay in ICU",
    y = "Length of Hospital Stay"
  )

## Plot using all ICU data -- lengths on same scale
## Good plot!!! 
ggplot(icu_data, aes(x = los_icu, y = los_hospital)) +
  geom_point() +
  labs(
    x = "Length of Stay in ICU, in days", 
    y = "Total Length of Stay in Hospital, in days" 
  )

## Is the relationship between ICU stay and hospital stay different with newborns removed? No
##icu_no_babies = icu_data %>% 
 ## filter(admission_type != "NEWBORN")

##ggplot(icu_no_babies, aes(x = los_icu, y = los_hospital)) +
  ##geom_point()


merged_data = merged_data %>% 
  mutate(length_proportion = (los_icu / los_hospital)) %>% 
  filter(!(length_proportion == "NA"))

icu_data %>% 
  ggplot(aes(x = los_hospital)) +
  geom_histogram() 

icu_data %>% 
  filter(los_hospital > 100) %>% 
  arrange(desc(los_hospital))


## How many proportions are greater than 1? (Too many)
## Limitation
icu_data %>% 
  mutate(prop_icu = los_icu/los_hospital) %>% 
  arrange(desc(prop_icu)) %>% 
  mutate(prop_true = ifelse(prop_icu > 1, 1, 0)) %>% 
  count(prop_true) %>% 
  filter(prop_true == 1)


## We ran into a challenge that some proportions of length of stay in ICU to length of stay in hospital (which includes stay in ICU) are greater than 1. This suggests possible flaws in the data. While we acknowledge that these flaws exist, for the purpose of this project we will filter out the proportions greater than 1 and analyze this subset of the dataset. 

## Filter out proportions greater than 1
true_prop = icu_data %>% 
  mutate(prop_icu = los_icu/los_hospital) %>% 
  filter(!(prop_icu > 1), !(prop_icu < 0))

filtered_admissions = admissions %>% 
  mutate(mortality = ifelse(is.na(deathtime), "No Mortality", "Mortality")) %>% 
  select(hadm_id, admission_type, diagnosis, religion, ethnicity, marital_status, insurance, mortality)

filtered_prop = left_join(true_prop, filtered_admissions, by = "hadm_id")
  
  
summary(filtered_prop$los_hospital)
summary(filtered_prop$los_icu)

## Proportion of time in ICU by admission type
## INCLUDE!!! 
ggplot(filtered_prop, aes(x = admission_type.x, y = prop_icu)) +
  geom_boxplot() +
  facet_grid(~mortality) +
  labs(
    x = "Admission Type", 
    y = "Proportion of Stay in ICU", 
    title = "Proportion of Time in ICU by Admission Type"
  )

## Proportion of time in ICU by insurance type
## Not significant
ggplot(filtered_prop, aes(x = insurance, y = prop_icu)) +
  geom_boxplot()

ggplot(filtered_prop, aes(x = insurance, y = prop_icu)) +
  geom_boxplot() 

length_of_care = lm(length_of_stay ~ admission_type + insurance + marital_status + religion + ethnicity + mortality + los_icu, data = merged_data)

AIC(length_of_care)

step(length_of_care, direction = "backward")

anova(lm(lm(length_of_stay ~ admission_type + insurance + marital_status + religion + ethnicity + + mortality + los_icu, data = merged_data)))
```


## Length of Stay Regression
```{r}
library(broom)

length_of_care = lm(length_of_stay ~ admission_type + insurance + marital_status + religion + ethnicity + mortality, data = difference)

## Why such a high p-value?
lm(mortality ~ length_of_stay, data = difference) %>% summary()

##fit1 = update(length_of_care, . ~ . -admission_type)
##summary(fit1)

##fit2 = update(length_of_care, . ~ . -insurance)
##summary(fit2)

##fit3 = update(length_of_care, . ~ . -marital_status)
##summary(fit3)

##fit4 = update(length_of_care, . ~ . -religion)
##summary(fit4)

##fit5 = update(length_of_care, . ~ . -ethnicity)
##summary(fit1)

##fit6 = update(length_of_care, . ~ . -mortality)
##summary(fit6)

##fit7 = update(length_of_care, . ~ . +diagnosis)
##summary(fit1)


summary(step(length_of_care, direction = "backward"))
```

