---
title: "length_of_care"
author: "Samantha Brown"
date: "11/17/2018"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(devtools)
library(readr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
admissions <- 
  read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names() %>% 
  mutate(diagnosis = factor(diagnosis))
```

## Exploratory analysis

```{r, output = FALSE}
n_admits = admissions %>% 
  nrow()

n_discharge = admissions %>% 
  filter(is.na(deathtime)) %>% 
  nrow()

round((1 - (n_discharge/n_admits)) * 100, digits = 2)
```

**Out of `r admissions %>% nrow()` patient admissions, `r admissions %>%  filter(is.na(deathtime)) %>%  nrow()` patients were ultimately discharged. The remaining `r round((1 - (n_discharge/n_admits)) * 100, digits = 2)`% were recorded deaths.**


```{r}
## Ethnicity
ethnicity = admissions %>% 
  group_by(ethnicity) %>% 
  count()
```

Need to do string replace to combine similar groups.

```{r}
## Married
marital_status = admissions %>% 
  group_by(marital_status) %>% 
  count() %>% 
  mutate(percent = n/nrow(admissions)*100) %>% 
  arrange(desc(percent))
```

`r marital_status[1,3]`% of the critical care patients were married, while `r marital_status[2, 3]`% of patients were single. The remaining `r round(100 - (marital_status[1,3] + marital_status[2, 3]), digits = 2)`% were either widowed, divorced, separated, had a life partner, or marital status was unknown.

```{r}
## Distribution of insurance type
admissions %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(percent_of_patients = round(n/nrow(admissions)*100, digits = 2)) %>% 
  arrange(desc(percent_of_patients)) %>% 
  select(-n)
```



```{r}
## Top 10 causes of recorded mortalities
admissions %>% 
  filter(deathtime > 0) %>%
  count(diagnosis) %>% 
  top_n(10) %>% 
  arrange(desc(n))
```

```{r}
## Look at sepsis diagnoses
sepsis = admissions %>% 
  filter(diagnosis == "SEPSIS") %>% 
  ## discharge = 1, mortality = 0
  mutate(mortality = ifelse(is.na(deathtime), 1, 0)) %>% 
  ## create indicator variable: married = 1, not married = 0
  mutate(marital_status = ifelse(marital_status == "MARRIED", 1, 0)) %>% 
  select(subject_id, insurance, marital_status, ethnicity, mortality)


## Look at top 10 causes of mortalities
top_10_causes = admissions %>% 
  filter(diagnosis == c("SEPSIS", "PNEUMONIA", "INTRACRANIAL HEMORRHAGE", "CONGESTIVE HEART FAILURE", "ALTERED MENTAL STATUS", "CARDIAC ARREST", "ABDOMINAL PAIN", "S/P FALL", "HYPOTENSION", "SUBARACHNOID HEMORRHAGE"))
```



```{r}
## Insurance -- mortality vs. no mortality 

## mortality
mortalities = admissions %>% 
  filter(deathtime > 0) %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(outcome = "mortality")

total_mortalities = sum(mortalities$n)

mortalities = mortalities %>% 
  mutate(patient_proportion = round(n/total_mortalities, digits = 4))

discharge = admissions %>% 
  filter(deathtime > 0) %>% 
  group_by(insurance) %>% 
  count() %>% 
  mutate(outcome = "no_mortality")

total_discharge = sum(discharge$n)

dishcharge = discharge %>% 
  mutate(patient_proportion = round(n/total_discharge, digits = 4)) %>% 
  rbind(mortalities) %>% 
  mutate(log_count = log(n)) %>% 
  ggplot(aes(x = reorder(insurance, log_count), y = log_count)) +
  geom_point(color = "blue") 
  
```


## Exploratory:
I want to discuss with you guys the challenge in finding length of care -- if someone was admitted on December 1, 1980 and discharged January 5, 1981, I'm not sure if we can just subtract the times? Using Francis's method, it results in negative number for month. Also, we cannot measure minutes/hours, because admit time does not have minutes

Should we measure length of stays by hours? then we can combine day and hour to one unit measurement

```{r}
library(lubridate)
admissions_data <- read_csv("./database/data/admissions.csv") %>% 
  janitor::clean_names()

```

```{r}
difference = admissions_data %>% 
  mutate(length_of_stay =  lubridate::as.duration(admittime %--% dischtime),
         mortality = ifelse(is.na(deathtime), 1, 0))

summary(difference$length_of_stay)

stay_in_days = difference %>% 
  mutate(length_of_stay = as.numeric(length_of_stay)/(60*60*24))


max(difference$length_of_stay)


max_length_in_days = max(as.numeric(difference$length_of_stay)/(60*60*24))
-1*min(as.numeric(difference$length_of_stay))/(60*60*24)

```



## Length of Stay Regression
```{r}
icu_data = read_csv("./database/data/icu_detail.csv")

library(broom)
length_of_care = lm(length_of_stay ~ admission_type + insurance + marital_status + religion + ethnicity + mortality, data = difference)

## Why such a high p-value?
lm(mortality ~ length_of_stay, data = difference) %>% summary()

##fit1 = update(length_of_care, . ~ . -admission_type)
##summary(fit1)

##fit2 = update(length_of_care, . ~ . -insurance)
##summary(fit2)

##fit3 = update(length_of_care, . ~ . -marital_status)
##summary(fit3)

##fit4 = update(length_of_care, . ~ . -religion)
##summary(fit4)

##fit5 = update(length_of_care, . ~ . -ethnicity)
##summary(fit1)

##fit6 = update(length_of_care, . ~ . -mortality)
##summary(fit6)

##fit7 = update(length_of_care, . ~ . +diagnosis)
##summary(fit1)


summary(step(length_of_care, direction = "backward"))

stay_vs_mortality = lm(length_of_stay ~ mortality, data = difference)
anova(stay_vs_mortality)
```

