---
title: "Medications"
author: "Laura Cosgrove"
date: "11/15/2018"
output: github_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
devtools::install_github("tidyverse/dplyr")
library(dplyr)
library(readr)
library(lubridate)
```
```

## MIMIC prescriptions description

```{r, cache=TRUE}

prescriptions_data <- read_csv("./database/data/PRESCRIPTIONS.csv.gz")
  
prescriptions_data = prescriptions_data %>% 
  janitor::clean_names() %>% 
  mutate(drug_type = factor(drug_type), dose_val_rx = as.integer(dose_val_rx),
          time_delta = as.duration(startdate %--% enddate),
          time_delta_num = as.numeric(time_delta))

```

What is the distribution of medications prescribed per person?
```{r}
#Plot distribution of total medications prescribed per person
prescriptions_data %>% 
  group_by(subject_id) %>% 
  add_tally() %>% 
  select(n, subject_id) %>% 
  distinct(subject_id, .keep_all = TRUE) %>% 
  ggplot(aes(x = n)) +
  geom_density()

#Plot distribution of distinct medications prescribed per person (with drug type not equal to an additive)

prescriptions_data %>%
  filter(drug_type != "ADDITIVE") %>% 
  group_by(subject_id) %>% 
  distinct(subject_id, drug, .keep_all = TRUE) %>% 
  add_tally() %>% 
  ggplot(aes(x = n)) +
  geom_density()
  

#What's the most common drug prescribed per person?



#Plot distribution of prescription length
prescriptions_data %>% 
  distinct(time_delta_num) %>% 
  ggplot(aes(x = time_delta_num)) +
  geom_density()

prescriptions_data %>% 
  arrange(time_delta) %>% 
  select(time_delta) 
  
prescriptions_data %>% 
  arrange(desc(time_delta)) %>% 
  select(time_delta) 
```

Since the data was collected only over ~11 years, we can safely eliminate some of the largest durations..

I'm having trouble right now because dplyr filter has some known issues with period and interval objects from lubridate. https://community.rstudio.com/t/dplyr-filter-issue-with-intervals-from-lubridate/9456/2. Hence all this weird code.


This is broken, will take another approach soon..
```{r}
prescriptions_data %>% 
  mutate(time_period = as.period(startdate %--% enddate),
        year_delta = year(time_period),
        year_delta_num = as.integer(year_delta)) %>% 
  filter(year_delta_num < 10 & year_delta_num < -10) %>% 
  select(time_period) %>% 
  arrange(time_period)
```

```{r}

```

