---
title: "Generating SAPS II scores"
author: "Laura Cosgrove"
date: "11/20/2018"
output: github_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(RPostgreSQL)
library(tidyverse)
library(dbplyr)
```


```{r dbconnect}
# Load configuration settings
dbdriver <- 'PostgreSQL'
host  <- '127.0.0.1'
port  <- '5432'
user  <- 'postgres'
password <- 'postgres'
dbname <- 'mimic'
schema <- 'mimiciii'
# Connect to the database using the configuration settings
con <- dbConnect(dbDriver(dbdriver), dbname = dbname, host = host, port = port, 
                 user = user, password = password)
# Set the default schema
dbExecute(con, paste("SET search_path TO ", schema, sep=" "))
```

Set this database as the connection for all future sql chunks:

```{r}
knitr::opts_chunk$set(connection = "con")
```

The above chunk works if you use knitr to generate the analysis! But in building the analysis it is maybe better to run your query by saving the sql query as a character object then using `dbGetQuery`.

Credit for SQL code authoring is not mine:

We need to generate some views into the database before the query to generate the severity scores works.
--  1) uofirstday - generated by urine-output-first-day.sql
--  2) ventdurations - generated by ventilation-durations.sql
--  3) vitalsfirstday - generated by vitals-first-day.sql
--  4) gcsfirstday - generated by gcs-first-day.sql
--  5) labsfirstday - generated by labs-first-day.sql
--  6) bloodgasarterialfirstday - generated by blood-gas-first-day-arterial.sql

```{r, eval = FALSE}
urine_view <- read_file("./database/mimic-code/concepts/firstday/urine-output-first-day.sql")
ventdurations_view <- read_file("./database/mimic-code/concepts/durations/ventilation-durations.sql")
vitals_view <- read_file("./database/mimic-code/concepts/firstday/vitals-first-day.sql")
gcs_view <- read_file("./database/mimic-code/concepts/firstday/gcs-first-day.sql")
labs_view <- read_file("./database/mimic-code/concepts/firstday/labs-first-day.sql")
bloodgasarterial_view <- read_file("./database/mimic-code/concepts/firstday/blood-gas-first-day-arterial.sql")
sapsii_view <- read_file("./database/mimic-code/concepts/severityscores/sapsii.sql")


#Generate materialized views
dbGetQuery(con, urine_view)
dbGetQuery(con, ventdurations_view)
dbGetQuery(con, vitals_view)
dbGetQuery(con, gcs_view)
dbGetQuery(con, labs_view)
dbGetQuery(con, bloodgasarterial_view)
dbGetQuery(con, sapsii_view)

```

Rewrite `sapsii_view` in R using `dbplyr`?

```{r}

```


```{r view}
#View sapsii_data
sapsii_query <- "SELECT *
              FROM sapsii i;"
sapsii_data <- as.tibble(dbGetQuery(con, sapsii_query))
write_csv(sapsii_data, path = "./database/sapsii.csv")

```

OMG Finally!

Plot distribution:

```{r}
sapsii_data %>% 
  ggplot(aes(x = sapsii)) + 
  geom_histogram()
```

A note in the SQL file is the following: 
-- Note:
--  The score is calculated for *all* ICU patients, with the assumption that the user will subselect appropriate ICUSTAY_IDs.
--  For example, the score is calculated for neonates, but it is likely inappropriate to actually use the score values for these patients.

```{sql connection=con}
SELECT *
FROM sapsii i
```



```{sql connection=con, eval=FALSE}
-- Calculate the AUROC of age for predicting in-hospital mortality
-- You can easily calculate the AUROC of any model you'd like by:
--  Replacing "PRED" with your predictor
--  Replacing "TAR" with the target (*must* be a binary target)

with datatable as (
select
  -- name the predictor "PRED"
  cast(adm.admittime as date) - cast(pat.dob as date) as PRED -- age is our predictor
  -- name the target variable "TAR"
  , case when adm.deathtime is not null then 1 else 0 end as TAR -- in-hospital mortality
from admissions adm
inner join patients pat
  on adm.subject_id = pat.subject_id
)
, datacs as (
select
  TAR
  -- calculate the cumulative sum of negative targets, then multiply by positive targets
  -- this has the effect of returning 0 for negative targets, and the # of negative targets below each positive target
  , TAR * SUM(1-TAR) OVER (ORDER BY PRED ASC ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW) AS AUROC
from datatable
)
select
  -- Calculate the AUROC as:
  --    SUM( number of negative targets below each positive target )
  -- /  number of possible negative/positive target pairs
  round(sum(AUROC) / (sum(TAR)*sum(1-TAR)),4) as AUROC
from datacs;

```

ROC curve for saps ii score:

```{r}
admissions <- read_csv("./database/data/ADMISSIONS.csv.gz") %>% 
  janitor::clean_names()
patients <- read_csv("./database/data/PATIENTS.csv.gz") %>% 
  janitor::clean_names()
sapsii <- read_csv("./database/sapsii.csv") %>% 
  janitor::clean_names()


admissions %>% 
  inner_join(., patients, by = "subject_id") %>% 
  filter(has_chartevents_data == 1) %>% 
  inner_join(., sapsii, by = "hadm_id") %>% 
  mutate(target = if_else(deathtime %in% NA, 0, 1),
         predictor = sapsii) %>%
  select(subject_id.x, target, predictor) %>% 
  group_by(predictor) %>% 
  summarize(deaths = sum(target),
            n = n()) %>% 
  mutate(frac_deaths = deaths/n) %>% 
  ggplot(aes(x = predictor, y = frac_deaths)) +
  geom_point(aes(color = n)) + 
  labs(x = "SAPS II Score",
       y = "Mortality Fraction", 
       title = "Predicting Mortality of ICU Patients with First-Day SAPS II scores") +
  theme_bw()
  

```

