---
title: "Medications"
author: "Laura Cosgrove"
date: "11/15/2018"
output: github_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
devtools::install_github("tidyverse/dplyr")
library(dplyr)
library(readr)
library(lubridate)
library(scales)

```
```

## MIMIC prescriptions description

```{r, cache=TRUE}

prescriptions_data <- read_csv("./database/data/PRESCRIPTIONS.csv.gz")
  
prescriptions_data = prescriptions_data %>% 
  janitor::clean_names() %>% 
  mutate(drug_type = factor(drug_type), dose_val_rx = as.integer(dose_val_rx),
          time_delta = lubridate::as.duration(startdate %--% enddate),
          time_delta_num = as.numeric(time_delta))

```

What is the distribution of medications prescribed per person?
```{r}
#Plot distribution of total medications prescribed per person
prescriptions_data %>% 
  group_by(subject_id) %>% 
  add_tally() %>% 
  select(n, subject_id) %>% 
  distinct(subject_id, .keep_all = TRUE) %>% 
  ggplot(aes(x = n)) +
  geom_density()

#Plot distribution of distinct medications prescribed per person (with drug type not equal to an additive)

prescriptions_data %>%
  filter(drug_type == "MAIN") %>% 
  group_by(subject_id) %>% 
  distinct(subject_id, drug, .keep_all = TRUE) %>% 
  add_tally() %>% 
  ggplot(aes(x = n)) +
  geom_density()
  

#What's the most commonly-administered drug per person?
prescriptions_data %>%
  filter(drug_type == "MAIN") %>% 
  group_by(drug) %>% 
  add_tally() %>% 
  distinct(drug, .keep_all = TRUE) %>% 
  select(n, drug, drug_type, formulary_drug_cd) %>% 
  arrange(desc(n)) %>% 
  head(n = 20L) %>% 
  knitr::kable()


#Are some drugs prescribed for longer periods of time compared to others?


#Plot distribution of prescription length
prescriptions_data %>% 
  distinct(time_delta_num) %>% 
  ggplot(aes(x = time_delta_num)) +
  geom_density()

prescriptions_data %>% 
  arrange(time_delta) %>% 
  select(time_delta) 
  
prescriptions_data %>% 
  arrange(desc(time_delta)) %>% 
  select(time_delta) 
```

Since the data was collected only over ~11 years, we can safely eliminate some of the largest durations..

I'm having trouble right now because dplyr filter has some known issues with period and interval objects from lubridate. https://community.rstudio.com/t/dplyr-filter-issue-with-intervals-from-lubridate/9456/2. Hence all this weird code.


This is broken, will take another approach soon.. Scratch that, it magically worked.
```{r}

prescriptions_data_time = prescriptions_data %>% 
  mutate(time_period = as.period(startdate %--% enddate),
        year_delta = year(time_period),
        year_delta_num = as.integer(year_delta)) %>% 
  filter(year_delta_num < 10 & year_delta_num > -10) %>% 
  select(time_delta, year_delta, startdate, enddate, drug, drug_type, formulary_drug_cd, subject_id)

#Look at if type of drugs are more commonly prescribed for larger amounts of time
#maybe a heatmap with drugs on an axis and time_delta on another axis, with instances prescribed as fill


#Look at if duration drug is prescribed changes over time

#prescriptions_data_time %>% 
# mutate(startdate = as.Date(startdate, "%m-%d-%y")) %>% 
# ggplot(aes(x = startdate, y = as.numeric(time_delta))) +
#    geom_point() +
#    scale_x_date(labels = date_format("%b/%y")) 
    

  
```

```{r other datasets, eval = FALSE}
#output_data <- read_csv("./database/data/OUTPUTEVENTS.csv.gz")
#caregivers_data = read_csv("./database/data/CAREGIVERS.csv.gz")
#cpt_events_data = read_csv("./database/data/CPTEVENTS.csv.gz")
#services_data = read_csv("./database/data/CPTEVENTS.csv.gz")
#transfers_data = read_csv("./database/data/TRANSFERS.csv.gz")


```

## Provide the number of medications dataset 

Join to admissions the number of medications prescribed per person. 
